---
title: "Robust TV: Huber and Tukey Data Terms"
author: "fmrismooth authors"
date: "`r format(Sys.Date())`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Robust TV: Huber and Tukey Data Terms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
library(fmrismooth)
```

Robust losses tolerate outliers better than the standard squared error. In fMRI, spikes or motion corruption can bias classical TV. The Huber loss behaves quadratically for small residuals and linearly for large ones; Tukey’s biweight further down‑weights extreme outliers.

`stv_robust4d()` implements robust data terms with space–time TV priors. Thresholds are derived from a noise estimate if not set explicitly.

## Parameters in practice

`lambda_s` and `lambda_t` are the TV weights as in standard TV. Select `loss = "huber"` or `loss = "tukey"`. The Huber threshold `delta` and Tukey cutoff `cthresh` determine the transition from inlier to outlier behavior; leaving them `NULL` estimates them from the data. The `alpha` parameter tunes the Tukey shape.

Temporal weights can reduce smoothing across high‑motion transitions: provide per‑frame motion parameters via `motion_params`, or pass explicit `temporal_weights`.

## Example

```{r}
d4 <- c(8, 8, 8, 10)
base <- array(100, dim = d4) + array(rnorm(prod(d4), sd = 5), dim = d4)
base[4,4,4,5] <- 220  # inject an outlier

rob_huber <- tv_robust4d(base, loss = "huber", iters = 20L)
rob_tukey <- tv_robust4d(base, loss = "tukey", iters = 20L)

c(var_huber = var(as.vector(rob_huber)),
  var_tukey = var(as.vector(rob_tukey)))
```

The robust losses reduce the influence of isolated spikes relative to standard TV. When artifacts correlate with motion, supplying `motion_params` improves temporal selectivity.

